Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-02-13 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
little update of main text

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(r)&=\alpha \, \frac{2^{1-\nu}}{\Gamma(\nu)}\left(\frac{\sqrt{2\nu}r}{\ell}\right)^{\!\nu} \! K_{\nu} \left(\frac{\sqrt{2\nu}r}{\ell}\right),
\end{align*}
%
where $\nu > 0$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell > 0$ and $\alpha > 0$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$, $\nu=1/2$, $\nu=3/2$ and $\nu=5/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp \left( -\frac{1}{2} \frac{r^2}{\ell^2}\right),  \\
k_{\frac{1}{2}}(r)&=\alpha \exp \left( -\frac{r}{\ell}\right),  \\
k_{\frac{3}{2}}(r)&=\alpha\left(1+\frac{\sqrt{3}r}{\ell}\right) \exp\! \left( -\frac{\sqrt{3}r}{\ell}\right),   \\
k_{\frac{5}{2}}(r)&=\alpha\left(1+\frac{\sqrt{5}r}{\ell} + \frac{5r^2}{3\ell^2}\right) \exp\! \left( -\frac{\sqrt{5}r}{\ell}\right). 
\end{align*}
%
The former is commonly known as the squared exponential or exponentiated quadratic covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form
%
\begin{align*}
k_{\infty}(r) = \alpha & \exp \left( -\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\right),\\
k_{\frac{1}{2}}(r) = \alpha & \exp \left( -\sum_{i=1}^{D}\frac{(x_i-x_i')}{\ell_i}\right),\\
k_{\frac{3}{2}}(r) = \alpha & \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right)\\
& \times \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right),\\
k_{\frac{5}{2}}(r) = \alpha & \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}+\frac{5}{3} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\;\right)\\
& \times \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{5(x_i-x_i')^2}{\ell_i^2}}\;\right).
\end{align*}
%
Notice that the previous expressions, $k_{\infty}(r)$, $k_{\frac{1}{2}}(r)$, $k_{\frac{3}{2}}(r)$ and $k_{\frac{5}{2}}(r)$, have been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. Using individual length-scales for each dimension turns the isotropic covariance function into a non-isotropic covariance function. That is, for a non-isotropic covariance function, the smoothness may vary across different input dimensions.

Stationary covariance functions can be represented in terms of their spectral densities \citep[see, e.g.,][]{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure \citep[\textit{Bochner's theorem}; see, e.g., ][]{akhiezer1993theory}. If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions are given by
%
\begin{align*}
S_{\nu}(\bm{\omega})&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)\, \ell^{2\nu}}\left(\frac{2\nu}{\ell^2}+4\pi^2\bm{\omega}^\intercal \bm{\omega} \right)^{-(\nu+D/2)}
\end{align*}
%
in $D$ dimensions, where vector $\bm{\omega}\in {\rm I\!R^D}$ is in the frequency domain, and $\ell$ and $\alpha$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular cases, where $\nu=\infty$, $\nu=1/2$, $\nu=3/2$ and $\nu=5/2$, take the form
%
\begin{align}
S_{\infty}(\bm{\omega})&= \alpha\,(\sqrt{2\pi})^D  \ell^D \exp(-\frac{1}{2}\ell^2 \bm{\omega}^\intercal \bm{\omega}), \label{eq_specdens_inf}  \\
S_{\frac{1}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+1}{2})}{\sqrt{\pi}\ell}\left(\frac{1}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+1}{2}} \label{eq_specdens_12},     \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2}){3}^{3/2}}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+3}{2}} \label{eq_specdens_32},     \\
S_{\frac{5}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+5}{2}){5}^{5/2}}{\frac{3}{4}\sqrt{\pi}\ell^5}\left(\frac{5}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+5}{2}}. \label{eq_specdens_52}   
\end{align}
%
For instance, with input dimensionality $D=3$ and $\bm{\omega}=(\omega_1,\omega_2,\omega_3)^\intercal$, the spectral densities written above take the form
%
\begin{align*}
S_{\infty}(\bm{\omega})&= \alpha \, (2\pi)^{3/2}  \prod_{i=1}^{3} \ell_i  \exp\!\left(-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 \omega_i^2 \right),   \\
S_{\frac{1}{2}}(\bm{\omega})&= \alpha \, 8\,\pi \prod_{i=1}^{3}\ell_i\left(1+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-2},    \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha \, 32\,\pi {3}^{3/2}\prod_{i=1}^{3}\ell_i\left(3+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-3},   \\
S_{\frac{5}{2}}(\bm{\omega})&= \alpha \, \frac{64}{3}\,\pi {5}^{5/2}\prod_{i=1}^{3}\ell_i\left(5+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-4}.
\end{align*}
%
where individual length-scales $\ell_i$ for each frequency dimension $\omega_i$ have been used.

@@

There is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. In this Section we discuss about the existing theoretical evidence that the minimum number of basis functions $m$ needed for an accurate approximation is linearly proportional to the the boundary factor $c$ and inversely proportional to the lengthscale $l$. In next Section \ref{subsec_empiric_relations} we conduct an empirical experiments to prove this expected linear proportionality relationship between $m$, $c$ and $l$ and compute the specific coefficients of proportionality for every kernel.

@@

Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_7_relationships} depict how these three factors interact and affect the accuracy of the HSGP approximation for a GP with square exponential covariance function and Mat\'ern ($\nu$=3/2) covariance function, respectively, and a single input dimension. More precisely, for a given GP model (with a squared exponential covariance function) with length-scale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_lscale_vs_J_vs_c} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). Similarly,  Figure \ref{fig6_7_relationships} shows the corresponding plot for the  Mat\'ern ($\nu$=3/2) covariance function. We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation*}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation*}
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:

\begin{itemize}
\item As $\ell$ increases, $m$ required for an accurate approximation decrease.
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve an accurate approximation.
\item For a given $\ell$ there exist a minimum $c$ under which a accurate approximation is never going to be achieved regardless of $m$. This fact can be seen in Figures~\ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_7_relationships} as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$). As $\ell$ increases, the minimum valid $c$ also increases.
\end{itemize}

@@

From the empirical observations the hypothesis of linear proportionality between $m$, $\frac{l}{S}$ (lenghscale $l$ relative to the half range of the input domain $S$) and $c$ has been proved. Following we show numerical functions governing these relationships for every kernel.

\paragraph*{Squared exponential}:
%
\begin{align}\label{eq_m_l_QE}
&m= 1.75 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 1.75 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_QE}
&c = c(l) \, \geq \, 3.2 \, \nicefrac{l}{S}         
\end{align}

\paragraph*{Mat\'ern ($\nu$=3/2)}:
%
\begin{align}\label{eq_m_l_mat32}
&m= 3.42 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 3.42 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat32}
&c \, \geq \, 4.5 \, \nicefrac{l}{S}         
\end{align}

\paragraph*{Mat\'ern ($\nu$=5/2)}:
%
\begin{align}\label{eq_m_l_mat52}
&m= 2.65 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 2.65 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat52}
&c \, \geq \, 4.1 \, \nicefrac{l}{S}         
\end{align}

Figures~\ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_7_relationships} were build for a GP with a unidimensional covariance function, which result in a surface depending on three variables, $m$, $c$ and $\ell$. An equivalent figure for a GP model with a two-dimensional covariance function would result in a surface depending on four variables, $m$, $c$, $\ell_1$ and $\ell_2$, which is more difficult to be graphically represented. More precisely, in the multi-dimensional case, whether the approximation is close enough might depend only on the ratio between wigglyness in every dimensions. For instance, in the two-dimensional case, it would depend on the ratio between $\ell_1$ and $\ell_2$ and could be graphically represented. Future research will focus on building useful graphs or analytical models that provide these relations in multi-dimensional cases. However, as an approximation, we can use the unidimensional GP conclusions in Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_7_relationships} to check the accuracy by analyzing individually the different dimensions of a multidimensional GP model.

@@

As stated above, Figures~\ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_7_relationships} provide the minimum length-scale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the length-scale $\ell$ controls the wigglyness of the function, it strongly influences the difficulty of obtaining accurate inference about the function from the data. Basically, if the length-scale estimate is accurate, we can expect the HSGP approximation to be accurate as well.

Having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum length-scale provided in Figure~\ref{fig6_lscale_vs_J_vs_c} or \ref{fig6_7_relationships} (depending on which kernel is used). If $\hat{\ell}$ exceeds this recommended minimum length-scale, the approximation is assumed to be good. If, however, $\hat{\ell}$ does not exceed recommended minimum length-scale, the approximation may be inaccurate and $m$ should be increased or $c$ decreased. We may also use this diagnostic in an iterative procedure by starting from some initial guess of $\ell$ and initial values for $m$ and $c$, and if the estimated $\hat{\ell}$ is below the minimum length-scale, repeat the process while increasing $m$ or decreasing $c$. As mentioned earlier, $c$ cannot be decreased too much as the lowest useful value of $c$ is restricted by the length-scale. Thus, increasing $m$ may usually the preferred approach.

%\begin{figure*}
%\centering
%\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{fig6_lscale_vs_J_vs_c.pdf}}
%\hspace{3mm}
%\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{fig6_lscale_vs_J_vs_c_zoomin.pdf}}
%\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), in the case of a square exponential covariance function. The right-side plot is a zoom in of the left-side plot.}
%  \label{fig7_lscale_vs_J_vs_c_Matern}
%\end{figure*}

%\begin{figure*}
%\centering
%\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{fig6_7_relationships.pdf}}
%\hspace{3mm}
%\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{fig7_lscale_vs_J_vs_c_zoomin_Matern.pdf}}
%\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), in the case of a Mat\'ern ($\nu$=3/2) covariance function. The right-side plot is a zoom in of the left-side plot.}
%  \label{fig6_7_relationships}
%\end{figure*}

\begin{figure*}
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig6_7_relationships.png}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), for the square exponential, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions.}
  \label{fig6_lscale_vs_J_vs_c}
\end{figure*}

@@

\paragraph*{Assumption of the diagnosis tool}: Under inaccurate HSGP approximation, the estimated lengthscale $\hat{l}$ is smaller than the exact GP lengthscale estimate $l$. Thus the diagnostic that determines is the approximation is sufficiently accurate is:
%
\begin{align}\label{eq_diagnostic}
&\hat{l} \pm 0.02 \geq l
\end{align}

>>>>

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(r)&=\alpha \, \frac{2^{1-\nu}}{\Gamma(\nu)}\left(\frac{\sqrt{2\nu}r}{\ell}\right)^{\!\nu} \! K_{\nu} \left(\frac{\sqrt{2\nu}r}{\ell}\right),
\end{align*}
%
where $\nu > 0$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell > 0$ and $\alpha > 0$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$, $\nu=3/2$ and $\nu=5/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp \left( -\frac{1}{2} \frac{r^2}{\ell^2}\right),  \\
%k_{\frac{1}{2}}(r)&=\alpha \exp \left( -\frac{r}{\ell}\right),  \\
k_{\frac{3}{2}}(r)&=\alpha\left(1+\frac{\sqrt{3}r}{\ell}\right) \exp\! \left( -\frac{\sqrt{3}r}{\ell}\right),   \\
k_{\frac{5}{2}}(r)&=\alpha\left(1+\frac{\sqrt{5}r}{\ell} + \frac{5r^2}{3\ell^2}\right) \exp\! \left( -\frac{\sqrt{5}r}{\ell}\right). 
\end{align*}
%
The former is commonly known as the squared exponential or exponentiated quadratic covariance function. As an example, assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernel $K_{\nu}$ written above takes the form
%
\begin{align*}
k_{\infty}(r) = \alpha & \exp \left( -\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\right).
%k_{\frac{1}{2}}(r) = \alpha & \exp \left( -\sum_{i=1}^{D}\frac{(x_i-x_i')}{\ell_i}\right),\\
%k_{\frac{3}{2}}(r) = \alpha & \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right)\\
%& \times \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right),\\
%k_{\frac{5}{2}}(r) = \alpha & \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}+\frac{5}{3} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\;\right)\\
%& \times \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{5(x_i-x_i')^2}{\ell_i^2}}\;\right).
\end{align*}
%
Notice that the previous expressions $k_{\infty}(r)$ has been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. Using individual length-scales for each dimension turns an isotropic covariance function into a non-isotropic covariance function. That is, for a non-isotropic covariance function, the smoothness may vary across different input dimensions.

Stationary covariance functions can be represented in terms of their spectral densities \citep[see, e.g.,][]{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure \citep[\textit{Bochner's theorem}; see, e.g., ][]{akhiezer1993theory}. If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions are given by
%
\begin{align*}
S_{\nu}(\bm{\omega})&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)\, \ell^{2\nu}}\left(\frac{2\nu}{\ell^2}+4\pi^2\bm{\omega}^\intercal \bm{\omega} \right)^{-(\nu+D/2)}
\end{align*}
%
in $D$ dimensions, where vector $\bm{\omega}\in {\rm I\!R^D}$ is in the frequency domain, and $\ell$ and $\alpha$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular cases, where $\nu=\infty$, $\nu=1/2$, $\nu=3/2$ and $\nu=5/2$, take the form
%
\begin{align}
S_{\infty}(\bm{\omega})&= \alpha\,(\sqrt{2\pi})^D  \ell^D \exp(-\frac{1}{2}\ell^2 \bm{\omega}^\intercal \bm{\omega}), \label{eq_specdens_inf}  \\
%S_{\frac{1}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+1}{2})}{\sqrt{\pi}\ell}\left(\frac{1}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+1}{2}} \label{eq_specdens_12},     \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2}){3}^{3/2}}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+3}{2}} \label{eq_specdens_32},     \\
S_{\frac{5}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+5}{2}){5}^{5/2}}{\frac{3}{4}\sqrt{\pi}\ell^5}\left(\frac{5}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+5}{2}}. \label{eq_specdens_52}   
\end{align}
%
For instance, with input dimensionality $D=3$ and $\bm{\omega}=(\omega_1,\omega_2,\omega_3)^\intercal$, the spectral densities written above take the form
%
\begin{align*}
S_{\infty}(\bm{\omega})&= \alpha \, (2\pi)^{3/2}  \prod_{i=1}^{3} \ell_i  \exp\!\left(-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 \omega_i^2 \right),   \\
%S_{\frac{1}{2}}(\bm{\omega})&= \alpha \, 8\,\pi \prod_{i=1}^{3}\ell_i\left(1+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-2},    \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha \, 32\,\pi {3}^{3/2}\prod_{i=1}^{3}\ell_i\left(3+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-3},   \\
S_{\frac{5}{2}}(\bm{\omega})&= \alpha \, \frac{64}{3}\,\pi {5}^{5/2}\prod_{i=1}^{3}\ell_i\left(5+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-4}.
\end{align*}
%
where individual length-scales $\ell_i$ for each frequency dimension $\omega_i$ have been used.

@@

There is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. In this section we discuss about the existing theoretical evidence that the minimum number of basis functions $m$ needed for an accurate approximation is linearly proportional to the the boundary factor $c$ and inversely proportional to the lengthscale $l$. In the next Section \ref{subsec_empiric_relations} we conduct an empirical analysis through simulations to prove this expected linear proportionality between $m$, $c$ and $l$ and compute the specific coefficients of proportionality for every kernel.

@@

\begin{figure*}
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig6_7_relationships.png}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), for the square exponential, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions.}
  \label{fig6_7_relationships}
\end{figure*}

Figure \ref{fig6_7_relationships} depicts how these three factors interact and affect the accuracy of the HSGP approximation for a GP with square exponential covariance function, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions and a single input dimension. More precisely, for a given GP model (with a particular covariance function) with length-scale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_7_relationships} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation*}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation*}
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:

\begin{itemize}
\item As $\ell$ increases, $m$ required for an accurate approximation decrease.
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve an accurate approximation.
\item For a given $\ell$ there exist a minimum $c$ under which a accurate approximation is never going to be achieved regardless of $m$. This fact can be seen in Figures~\ref{fig6_7_relationships} and \ref{fig6_7_relationships} as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$). As $\ell$ increases, the minimum valid $c$ also increases.
\end{itemize}

@@

From the empirical observations the hypothesis of linear proportionality between $m$, $\frac{l}{S}$ (lenghscale $l$ relative to the half range of the input domain $S$) and $c$ has been proved. Following we show the numerical functions governing these relationships for every kernel.

\paragraph*{Squared exponential}:
%
\begin{align}\label{eq_m_l_QE}
&m= 1.75 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 1.75 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_QE}
&c \geq \, 3.2 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2      
\end{align}

\paragraph*{Mat\'ern ($\nu$=3/2)}:
%
\begin{align}\label{eq_m_l_mat32}
&m= 3.42 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 3.42 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat32}
&c \, \geq \, 4.5 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2        
\end{align}

\paragraph*{Mat\'ern ($\nu$=5/2)}:
%
\begin{align}\label{eq_m_l_mat52}
&m= 2.65 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 2.65 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat52}
&c \, \geq \, 4.1 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2       
\end{align}

Notice that we establish a lower bound for $c$ set to 1.2.

Figure~\ref{fig6_7_relationships} was build for a GP with a unidimensional covariance function, which result in a surface depending on three variables, $m$, $c$ and $\ell$. An equivalent figure for a GP model with a two-dimensional covariance function would result in a surface depending on four variables, $m$, $c$, $\ell_1$ and $\ell_2$, which is more difficult to be graphically represented. More precisely, in the multi-dimensional case, whether the approximation is close enough might depend only on the ratio between wigglyness in every dimensions. For instance, in the two-dimensional case, it would depend on the ratio between $\ell_1$ and $\ell_2$ and could be graphically represented. Future research will focus on building useful graphs or analytical models that provide these relations in multi-dimensional cases. However, as an approximation, we can use the unidimensional GP conclusions in Figures \ref{fig6_7_relationships} and \ref{fig6_7_relationships} to check the accuracy by analyzing individually the different dimensions of a multidimensional GP model.

@@

As stated above, Figure~\ref{fig6_7_relationships} provides the minimum length-scale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the length-scale $\ell$ controls the wigglyness of the function, it strongly influences the difficulty of obtaining accurate inference about the function from the data. Basically, if the length-scale estimate is accurate, we can expect the HSGP approximation to be accurate as well.

Having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum length-scale provided in Figure~\ref{fig6_7_relationships} (depending on which kernel is used). If $\hat{\ell}$ exceeds this recommended minimum length-scale, the approximation is assumed to be good. If, however, $\hat{\ell}$ does not exceed recommended minimum length-scale, the approximation may be inaccurate and $m$ should be increased or $c$ decreased. We may also use this diagnostic in an iterative procedure by starting from some initial guess of $\ell$ and initial values for $m$ and $c$, and if the estimated $\hat{\ell}$ is below the minimum length-scale, repeat the process while increasing $m$ or decreasing $c$. As mentioned earlier, $c$ cannot be decreased too much as the lowest useful value of $c$ is restricted by the length-scale. Thus, increasing $m$ may usually the preferred approach.

@@

\paragraph*{Assumption of the diagnosis tool}: Under inaccurate HSGP approximation, the estimated lengthscale $\hat{l}$ is smaller than the minimum length-scale $l^*$ determined by $m$ and $c$ in Figure~\ref{fig6_7_relationships}. Thus the diagnostic pointing towards an accurate approximation is:
%
\begin{align}\label{eq_diagnostic}
&\hat{l} \pm 0.02 \geq l^*
\end{align}
